# GO ONMF

## Introduction

Kim and Yu, A Mutation Profile for Top-k Patient Search Exploiting Gene-Ontology
and Orthogonal Non-negative Matrix Factorization, Bioinformatics, 2015,
http://dx.doi.org/10.1093/bioinformatics/btv409

Kim and Yu recently published the above paper in which they develop a method
for creating a mutation-gene-ontology profile. As part of the overall procedure, 
they described (Section 2.3, *Gene Function Profile*):

> To reduce the term correlation, we use only the most specific terms, i.e. the leaf 
> node term after propagating the scores of the non-leaf terms down to the leaf node 
> terms. This approach also resolves the problem of evaluating genes annotated with 
> general term as the effect of the gene of function identification is spread out over 
> several leaf node terms.

This sounds troubling to my ears, because the Gene Ontology (GO) is a *directed 
acyclic graph* (DAG), wherein there are specific relationships between the terms, and
there is a directionality. There is also an expectation that when a gene product
is annotated to a *specific* term, then that gene product is also automatically
annotated to all of the *less-specific* **parent** terms in the DAG. 

What the procedure quoted above sounds like it is doing is essentially **adding**
more specific gene product to GO annotations, which as far as I know is not allowed.

This document is a record of my investigation as to whether *more specific* gene
product - GO annotations are generated during the creation of the mutation 
profiles.

## Software

Kim and Yu made an Octave version of their software available at https://sites.google.com/site/postechdm/research/implementation/orgos, and a
tarball of the Octave version can be downloaded from https://sites.google.com/site/postechdm/research/implementation/orgos/ONMF_octave-simple.zip

```
wget https://sites.google.com/site/postechdm/research/implementation/orgos/ONMF_octave-simple.zip
mkdir ONMF_source
unzip ONMF_octave-simple.zip -d ONMF_source
```

As part of the software package, a few data files are provided. The ones that
are important for our purposes include: 

* ONMF_source/brca/go_(merged).csv - the index to GO term file
* ONMF_source/brca/gene_(merged).csv - the index to gene file
* exp_onmf_brca.m - the file that runs the *brca* analysis
* network_gene2go(merged).csv - the gene 2 GO annotations

## Running Code

After modifying *exp_onmf_brca.m* with:

> library_path = '/home/rmflight/Projects/personal/onmf/ONMF_source/ONMF_octave';

And in `Octave` I ran:

```
cd brca
run exp_onmf_brca.m
```

As this runs, it generates *gene2go.mat*, which is the propogated scores of gene
to GO associations.

We will use `Octave` to generate some indices and print them to files that can be
double checked in `R` (see ONMF_source/brca/exp_onmf_brca.m)

```{r print_octave_code, echo = FALSE, comment=''}
# from run_onmf_brca.m
octave_code <- scan("ONMF_source/brca/exp_onmf_brca_mod.m", sep = "\n", what = "character")
cat(octave_code, sep = "\n")
```

## Prerequisites

To continue this analysis, you will need some other `R` packages installed.

```{r setup, eval=FALSE}
install.packages(c("readr", "dplyr"))
library(BiocInstaller)
biocLite(c("GO.db", "graph", "org.Hs.eg.db"))
```

```{r load_packages, message=FALSE}
library(readr)
library(dplyr)
library(GO.db)
library(org.Hs.eg.db)
```


## Compare Original and Propagated

Now that we have the original and propagated indices in a format that we should
be able to read, we can compare them and check how things are being propagated,
whether to more or less specific terms.

### Read Indices

```{r read_indices}
org_indices <- scan(file = "ONMF_source/brca/org_notzero", what = numeric(), 
                    sep = " ")
org_indices <- org_indices[!is.na(org_indices)]
prop_indices <- scan(file = "ONMF_source/brca/prop_notzero", what = numeric(),
                     sep = " ")
prop_indices <- prop_indices[!is.na(prop_indices)]
```

### Read Other Data Files

```{r read_other}
go_map <- read_csv("ONMF_source/brca/go_(merged).csv", col_names = c("loc", "GO"))
gene_map <- read_csv("ONMF_source/brca/gene_(merged).csv", col_names = c("loc", "GENE"))
```

### Actually Compare

First we need to actually get out the GO id's.

```{r go_ids}
org_go <- dplyr::filter(go_map, loc %in% org_indices) %>% dplyr::select(GO) %>% 
  unlist(., use.names = FALSE) %>% unique()
prop_go <- dplyr::filter(go_map, loc %in% prop_indices) %>% dplyr::select(GO) %>% 
  unlist(., use.names = FALSE) %>% unique()
```

And then what gene are we supposedly working with? I randomly chose the one with
the index of **783**.

```{r gene_id}
gene_id <- dplyr::filter(gene_map, loc %in% 783) %>% dplyr::select(GENE) %>% 
  unlist(., use.names = FALSE)
```

The gene name is `r gene_id`.

Now, are the terms in `org_go` parents of the terms in `prop_go`? We will simply
make a calls to `GOBPANCESTORS` and save the results.

```{r get_ancestors}
# first check that everything is BP
org_terms <- GOTERM[org_go]
unique(Ontology(org_terms))

prop_terms <- GOTERM[prop_go]
unique(Ontology(prop_terms))

# get ancestors
prop_ancestors <- mget(prop_go, GOBPANCESTOR, ifnotfound = NA) %>% 
  unlist(., use.names = FALSE) %>% unique()

sum(org_go %in% prop_ancestors)
```

Hmmm, there are `r sum(org_go %in% prop_ancestors)` of the originals in the
propagated. This seems not right. Maybe they are children instead?

```{r get_children}
org_ancestors <- mget(org_go, GOBPANCESTOR, ifnotfound = NA) %>%
  unlist(., use.names = FALSE) %>% unique()

sum(prop_go %in% org_ancestors)
```

OK, this doesn't look right either. 

## Check GO Relationships

There is a file that has the distances between GO terms, encoded as:

* GO term 1 | distance | GO term 2

For adjacent terms, it appears that the distance is 0.5. So lets extract all
of them, and compare them to the data in `GO.db`.

With this gene id, we can lookup the GO annotations for the gene stored in
`org.Hs.eg.db`.

```{r get_go}
gene_go <- AnnotationDbi::select(org.Hs.eg.db, keys = gene_id, columns = "GO",
                                 keytype = "SYMBOL")
gene_allgo <- AnnotationDbi::select(org.Hs.eg.db, keys = gene_id, columns = "GOALL",
                                    keytype = "SYMBOL")
```

And lookup how many of the ONMF GO terms are in these two lists.

```{r compare_onmf_to_official}

```

