---
title: "RMF Investigation of GO-ONMF"
author: "Robert M Flight <rflight79@gmail.com>"
date: "`r Sys.time()`"
output: md_document
---

# GO ONMF

## Contact with Authors

As of Dec 1, 2015, a copy of this report was sent to Hwanjo Yu as the corresponding
author listed on the manuscript for feedback, either in email, or ideally as issues
or pull requests on this `github` repo.

## Introduction

Kim, Sael and Yu, "A Mutation Profile for Top-k Patient Search Exploiting Gene-Ontology
and Orthogonal Non-negative Matrix Factorization", Bioinformatics, 2015,
http://dx.doi.org/10.1093/bioinformatics/btv409

Kim, Sael and Yu recently published the above paper in which they develop a method
for creating a mutation-gene-ontology profile. As part of the overall procedure, 
they described (Section 2.3, *Gene Function Profile*):

> To reduce the term correlation, we use only the most specific terms, i.e. the leaf 
> node term after propagating the scores of the non-leaf terms down to the leaf node 
> terms. This approach also resolves the problem of evaluating genes annotated with 
> general term as the effect of the gene of function identification is spread out over 
> several leaf node terms.

This sounds troubling to my ears, because the Gene Ontology (GO) is a *directed 
acyclic graph* (DAG), wherein there are specific relationships between the terms, and
there is a directionality. There is also an expectation that when a gene product
is annotated to a *specific* term, then that gene product is also automatically
annotated to all of the *less-specific* **parent** terms in the DAG. 

What the procedure quoted above sounds like it is doing is essentially **adding**
**more specific** GO annotations to a gene product, which as far as I know is not allowed.

From [The Gene Ontology Consortium, 2001](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC311077/):

> The pathway from a child term to its top-level parent(s) must always be true.

This is why when considering gene-GO annotations, we consider the specific term
annotation, and **all** parent terms as well, because of the **true path** rule.
However, the converse is not true.

This document is a record of my investigation as to whether **more specific** GO
annotations are being added to gene products during the creation of the mutation 
profiles.

## Software

Kim and Yu made an `Octave` version of their software available at https://sites.google.com/site/postechdm/research/implementation/orgos, and a
tarball of the Octave version can be downloaded from https://sites.google.com/site/postechdm/research/implementation/orgos/ONMF_octave-simple.zip

This code was used to fetch a copy of the software:

```
wget https://sites.google.com/site/postechdm/research/implementation/orgos/ONMF_octave-simple.zip
mkdir ONMF_source
unzip ONMF_octave-simple.zip -d ONMF_source
```

As part of the software package, a few data files are provided. The ones that
are important for our purposes include: 

* ONMF_source/brca/go_(merged).csv - the index to GO term file
* ONMF_source/brca/gene_(merged).csv - the index to gene file
* exp_onmf_brca.m - the file that runs the *brca* analysis
* network_gene2go(merged).csv - the gene 2 GO annotations

## Running Code

After modifying *exp_onmf_brca.m* to have the correct library path, and then
adding some code that will save **gene - GO** indices before and after the
optimization step (see below for the actual code run), I `cd` to the `ONMF_source`
directory, and then start `Octave`, and then run:

```
cd brca
run exp_onmf_brca_mod.m
```

As this runs, it generates *gene2go.mat*, which is the propagated scores of gene
to GO associations, and then saves indices (see [file](ONMF_source/brca/exp_onmf_brca_mod.m)):

```{r print_octave_code, echo = FALSE, comment=''}
# from run_onmf_brca.m
octave_code <- scan("ONMF_source/brca/exp_onmf_brca_mod.m", sep = "\n", what = "character")
cat(octave_code, sep = "\n")
```

## Prerequisites

To continue this analysis, you will need some other `R` packages installed.

```{r setup, eval=FALSE}
install.packages(c("readr", "dplyr"))
library(BiocInstaller)
biocLite(c("GO.db", "graph", "org.Hs.eg.db"))
```

## Load Packages

```{r load_packages, message=FALSE}
library(readr)
library(dplyr)
library(GO.db)
library(org.Hs.eg.db)
```

## Check Actual Input Data

Lets check that the actual data being used as input looks right. **Note** that 
1 gets added to the map indices so that we easily work with the data from the
matrices in `Octave`. The input data includes the `go_map`, that maps GO terms
to indices, `gene_map` mapping gene symbols to indices, `gene2go` mapping the
GO annotations to genes, and `go2go`, giving the parent-child relationships of
the GO terms.

```{r load_annotation}
go_map <- read_csv("ONMF_source/brca/go_(merged).csv", 
                   col_names = c("loc", "GO"))
go_map$loc <- go_map$loc + 1
gene_map <- read_csv("ONMF_source/brca/gene_(merged).csv", 
                     col_names = c("loc", "GENE"))
gene_map$loc <- gene_map$loc + 1
gene2go <- read_csv("ONMF_source/brca/network_gene2go(merged).csv",
                    col_names = c("gene", "score", "GO"))
gene2go$gene <- gene2go$gene + 1
gene2go$GO <- gene2go$GO + 1
go2go <- read_csv("ONMF_source/brca/network_allgo2allgo(merged).csv",
                  col_names = c("GO1", "distance", "GO2"))
go2go$GO1 <- go2go$GO1 + 1
go2go$GO2 <- go2go$GO2 + 1
```

And add the actual gene symbols and GO IDs to the `gene2go` to enable easy comparisons
with any other data source we can find.

```{r translate_annotation}
gene2go$geneid <- unlist(gene_map[match(gene2go$gene, gene_map$loc), "GENE"],
                         use.names = FALSE)
gene2go$goid <- unlist(go_map[match(gene2go$GO, go_map$loc), "GO"],
                       use.names = FALSE)
```

And then query `org.Hs.eg.db` for the gene-annotations as well.

```{r gene_annotation_from_source}
hs_gene2go <- select(org.Hs.eg.db, keys = unique(gene2go$geneid),
                     columns = c("GO", "ONTOLOGY"), keytype = "SYMBOL")
hs_gene2go <- dplyr::filter(hs_gene2go, ONTOLOGY == "BP", EVIDENCE != "IEA")
```

### Compare Gene-GO Annotations

And compare the annotations for the genes between the data from Kim and from
the Bioconductor database.

```{r compare_annotation_sources}
annotation_overlap <- vapply(unique(hs_gene2go$SYMBOL), function(in_loc){
  hs_data <- dplyr::filter(hs_gene2go, SYMBOL == in_loc) %>%
    dplyr::select(., GO) %>% unlist()
  onmf_data <- dplyr::filter(gene2go, geneid == in_loc) %>%
    dplyr::select(., goid) %>% unlist()
  sum(hs_data %in% onmf_data) / length(hs_data)
}, numeric(1))
```

And check the distribution of overlap:

```{r distribution_overlap}
hist(annotation_overlap)
```

That is great, over 4000 of the genes have what appear to be very similar terms!

### Compare GO-GO Relationships

Lets also check the **GO2GO** data.

```{r check_go2go}
go_notself <- dplyr::filter(go2go, distance != 1)
go_notself$ID1 <- unlist(go_map[match(go_notself$GO1, go_map$loc), "GO"], use.names = FALSE)
go_notself$ID2 <- unlist(go_map[match(go_notself$GO2, go_map$loc), "GO"], use.names = FALSE)

onmf_pc <- split(go_notself$ID2, go_notself$ID1)
go_pc <- mget(names(onmf_pc), GOBPCHILDREN, ifnotfound = NA)

go2go_overlap <- vapply(names(onmf_pc), function(in_loc){
  go_data <- go_pc[[in_loc]]
  onmf_data <- onmf_pc[[in_loc]]
  sum(go_data %in% onmf_data) / length(go_data)
}, numeric(1))
```

And plot it:

```{r plot_go2go_comparison}
hist(go2go_overlap)
```

## Work With Single Gene (A2M)

Now lets work with a single gene, and see what is going on. We'll keep it simple
and play with the first one in the list, *A2M*.

### Grab Starting and Propagated Terms

```{r a2m_data}
onmf_a2m_start <- dplyr::filter(gene2go, geneid == "A2M") %>%
  dplyr::select(goid) %>% unlist(., use.names = FALSE) %>% sort()

onmf_a2m_indices_org <- scan(file = "ONMF_source/brca/org_notzero", what = numeric(),
                         sep = " ")
onmf_a2m_indices_org <- onmf_a2m_indices_org[!is.na(onmf_a2m_indices_org)]
onmf_a2m_matrix_org <- dplyr::filter(go_map, loc %in% onmf_a2m_indices_org) %>%
  dplyr::select(GO) %>% unlist(., use.names = FALSE) %>% sort()

onmf_a2m_indices_prop <- scan(file = "ONMF_source/brca/prop_notzero", 
                              what = numeric(), sep = " ")
onmf_a2m_indices_prop <- onmf_a2m_indices_prop[!is.na(onmf_a2m_indices_prop)]
onmf_a2m_matrix_prop <- dplyr::filter(go_map, loc %in% onmf_a2m_indices_prop) %>%
  dplyr::select(GO) %>% unlist(., use.names = FALSE) %>% sort()
```

### Compare To Database

Now, how many of the terms in the starting gene annotation actually in the database
annotation?

```{r check_start_annotation}
hs_a2m_go <- dplyr::filter(hs_gene2go, SYMBOL == "A2M") %>%
  dplyr::select(GO) %>% unlist(., use.names = FALSE) %>% unique()
sum(onmf_a2m_matrix_org %in% hs_a2m_go) / length(onmf_a2m_matrix_org)
```

OK, seems we have most of them. Good.

### Check if Propagated Are Children of Original

Next, are the propagated terms actually children of the starting terms? We will
use multiple iterations of calls to `GOBPCHILDREN` and see if we get progressively
more of the terms.

```{r get_children}
n_iteration <- 10
query_list <- onmf_a2m_matrix_org
children_overlap <- numeric(n_iteration + 1)

for (i_iter in seq(1, n_iteration)) {
  children_overlap[i_iter + 1] <- 
    sum(onmf_a2m_matrix_prop %in% query_list) / length(onmf_a2m_indices_prop)
  
  query_child <- mget(query_list, GOBPCHILDREN, ifnotfound = NA) %>% 
    unlist(., use.names = FALSE) %>% unique()
  query_child <- query_child[!is.na(query_child)]
  query_list <- unique(c(query_list, 
                       query_child))
}
```

```{r plot_it}
plot(seq(0, n_iteration), children_overlap)
```

**FOUND THEM**. The propagated terms have **NO** overlap with the original
terms (see 0's at the start of the graph), and are children of the original terms. 
Just to be sure, I will also compare the propagated terms to the `GOALL` annotation,
which would be all of the GO terms that can **technically** be considered as
**annotated** to the gene (see *true path rule* from the Introduction), 
in that all **terms** and their **parent** terms are **annotated** to a gene product.

```{r get_all}
hs_a2m_all <- select(org.Hs.eg.db, keys = "A2M", columns = "GOALL", 
                     keytype = "SYMBOL")
head(hs_a2m_all)
sum(hs_a2m_all$GOALL %in% onmf_a2m_matrix_org) # the starting annotation
sum(hs_a2m_all$GOALL %in% onmf_a2m_matrix_prop) # propogated annotation
```

From this, none of the **propogated** terms can be properly considered as 
annotations to **A2M**. Effectively, the code has generated a **new set** of
GO annotations to **A2M**. I am sure this is the situation for all of the
genes.

## Back to Parent

On Dec 3, Sael replied to my email (see [correspondence](correspondence/sael_from_2015-12-03.md)),
and the crux of his response amounts to propagating back up to parents to get
significant / interpretable GO terms (also in the supplemental materials, I have
added a [PDF of the supplementary materials](supp_data/SUPPLEMENTARY_DATA.pdf). 
This seems rather odd to end up doing two sets of propagation, one down to leaf
terms to make the ONMF better (I assume), and then another up to parent terms 
to get more interpretable / significant results.

I am also concerned that propagating back up will actually result in more terms
than were originally annotated to a gene of interest. This is possible because
of the DAG nature of the GO and multi-parent / multi-child relationships.

Lets actually check if a round of propagating down to children and then back up
to parents results in the same or different set of parent terms than was started 
with. I will use the same set of terms that were annotated to *A2M* above.

I am going to use a set number of iterations of propagation (10) both down and
up to demonstrate the possible problem.

```{r prop_down}
children_down <- list(11)
children_down[[1]] <- hs_a2m_go

for (i_iter in seq(2, 11)) {
  tmp_down <- mget(children_down[[i_iter - 1]], GOBPCHILDREN, ifnotfound = NA)
  na_entries <- sapply(tmp_down, function(x){sum(is.na(x)) == 1})
  all_children <- c(unlist(tmp_down, use.names = FALSE), names(na_entries)[na_entries])
  children_down[[i_iter]] <- all_children[!is.na(all_children)]
}

len_children <- vapply(children_down, length, numeric(1))
len_children
```

From this, it looks like all of the children are captured by step *5* of the
propagation. So it will be 5 steps of propagation back up to parents.

```{r prop_up}
parents_up <- list(6)
parents_up[[1]] <- children_down[[6]]

for (i_iter in seq(2, 6)) {
  tmp_up <- mget(parents_up[[i_iter - 1]], GOBPPARENTS, ifnotfound = NA)
  na_entries <- vapply(tmp_up, function(x){sum(is.na(x)) == 1}, logical(1))
  all_parents <- c(unlist(tmp_up, use.names = FALSE), names(na_entries)[na_entries])
  parents_up[[i_iter]] <- all_parents[!is.na(all_parents)]
}
len_parents <- vapply(parents_up, length, numeric(1))
len_parents
```

As I suspected, going from the propagated children back up to parents (even a 
limited number of iterations) results in a huge number of terms, many more than
was in the initial list. Granted, this did not seem to happen to such a degree
in the manuscript, because they were able to get significant enrichment of
parent terms, so only a limited subset had to have been returned by the iterative
matrix multiplication. 

# Concluding Thoughts

However, there is nothing in the manuscript that demonstrates that the initial
propagation to child terms is necessary for performance of the method, or that
using a GOSlim does not work, or that the same / similar parent terms are returned
by propagating back up. I am surprised that the reviewers did not ask for any
of these things given the nature of the data.
